---
title: 'From CliFlo to clifro: An Introduction'
author: "Blake Seers"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 2
    highlight: haddock
    theme: readable
---

<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Introduction to clifro}
-->

```{r, echo=FALSE}
library(clifro)
```

# Introduction
The National Climate Database holds climate data from around 6,500 climate 
stations around New Zealand including some offshore and Pacific Islands. Over
600 stations are currently active and are still receiving valuable climate data.
[CliFlo](http://cliflo.niwa.co.nz/) is a web interface to the database managed 
by [NIWA](http://www.niwa.co.nz/), allowing users to submit queries and retrieve 
ten-minute, hourly, daily or summary data. The use of CliFlo is free given that 
the user has [subscribed](http://cliflo.niwa.co.nz/pls/niwp/wsubform.intro) and 
accepted NIWA's [terms and conditions](http://cliflo.niwa.co.nz/doc/terms.html).

The `clifro` package is designed to make CliFlo queries much simpler and
provide extensions that are currenly not offered by CliFlo. The intention is 
to simplify the data extraction, manipulation, exploration and visualisation
processes and easily create publication-worthy graphics for some of the primary 
datatypes, especially for users with limited or non-existent previous R 
experience. Experienced useRs will also find this package helpful for maximising
efficiency of climate data integration with R for further analysis, modelling or 
export.

This vignette provides an introduction to the `clifro` package demonstrating the
primary functionality by way of example. For more information on any of the 
functions in the `clifro` package the user is referred to the help index for
the `clifro` package, `help(package = "clifro")`.

# Create a clifro User
As stated above, if the intention is to extract data from any station other than
Reefton Ews (subscription-free) and to maximise the potential of `clifro`, a 
valid [subscription](http://cliflo.niwa.co.nz/pls/niwp/wsubform.intro) is 
needed.

The `cf.user` function is all that is required to create a valid `clifro` user,

```{r,eval=FALSE}
me = cf.user("username", "password")
```

where `username` and `password` is substituted for the user's CliFlo 
credentials.

# Create clifro Datatypes

Once the user has been authenticated, the next step is to choose the datatypes
of interest, see the [choose datatypes vignette][chooseDatatype] for details on
choosing datatypes. For this example we are interested in daily MSL atmospheric 
pressure, minimum and maximum temperature extremes (deg C), daily rainfall (mm)
and daily surface wind.
(m/s).

```{r}
my.dts = cf.datatype(select_1 =     c(7,  4,  3,  2), 
                     select_2 =     c(1,  2,  1,  1), 
                     check_box = list(3,  1,  1,  4), 
                     combo_box =    c(NA, NA, NA, 1))
my.dts
```

# Create clifro Stations

The third requisite for a valid `clifro` query is the station where the
data has been collected. If the agent numbers of the required CliFlo stations
are known, the only function needed to create a clifro station `cfStation` 
object is `cf.station`. See the [choose station vignette][chooseStation]
for help with choosing stations when the agent numbers are unknown, and the 
[working with clifro stations vignette][clifrostation] for further information
and methods on `cfStation` objects.

For this example we are interested in assessing how these datatypes differ in
various parts of the country by taking a selection of stations from various
regions. These include a station from Invercargill (5814), Nelson (4241), 
Hamilton (2112) and Auckland (1962)

```{r}
my.stations = cf.station(5814, 4241, 2112, 1962)
my.stations[, 1:5]
```

# Retrieve the CliFlo Data

Now that we have a valid `clifro` user and the datatypes and stations of 
interest, a `clifro` query can be conducted using the `cf.query` function. We
are interested in all available data from 2012 to 2014.

```{r,eval=FALSE}
cf.datalist = cf.query(user = me, 
                       datatype = my.dts, 
                       station = my.stations, 
                       start_date = "2012-01-01 00", 
                       end_date = "2014-01-01 00")
cf.datalist
```

```
## List containing clifro data frames:
##               data      type              start                end rows
## df 1)     Pressure  9am only (2012-01-01  8:00) (2013-01-01  8:00) 1468
## df 2)      Max_min     Daily (2012-01-01  8:00) (2013-12-31  8:00) 2923
## df 3)         Rain     Daily (2012-01-01  8:00) (2013-12-31  8:00) 2923
## df 4) Surface Wind  9am only (2012-01-01  8:00) (2013-01-01  8:00) 1468
```

We can see that the pressure and surface wind data only span one year.

# Plot the CliFlo Data

There is now a list of 4 dataframes in R containing all the available data for
each of the stations and datatypes chosen above. The plotting is simply done 
with a call to `plot`, the type of plot and plotting options depends on the 
datatype. See `?'plot,cfData-method'` for more details. The following are
examples of some of the plots possible with `clifro`, note how the optional 
`ggtheme`argument changes the look of the plots.

## MSL Atmospheric Pressure

```{r,eval=FALSE}
plot(cf.datalist, which = 1, ggtheme = "bw")

# Equivalent to 
# plot(cf.datalist[1], ggtheme = "bw")
```

![atmospheric pressure][mslAtmosPress]

The `which` argument tells `clifro` which of the four dataframes to plot. Since 
the plot method returns a `ggplot` object, we can easily modify the plots using 
[ggplot2](http://ggplot2.org/).

```{r,eval=FALSE}
# Load the ggplot2 library
library(ggplot2)

# Add a loess smoother with a span of a third of the window
plot(cf.datalist, which = 1, ggtheme = "bw",
     strip.text = element_text(size = 14)) + 
  geom_smooth(method = "loess", span = 1/3)
```

![atmospheric pressure][mslAtmosPress2]

## Daily Temperature Extremes

```{r,eval=FALSE}
plot(cf.datalist, which = 2, ggtheme = "linedraw")
```

![temperature extremes][temperature]

## Rain

Currently there are two possible default plots available for rainfall; with or
without soil deficit/runoff.

```{r,eval=FALSE}
plot(cf.datalist, which = 3, ggtheme = "light")
```

![rain and runoff][rainRunoff]

```{r, eval=FALSE}
# Don't plot the soil deficit and runoff
plot(cf.datalist, which = 3, include_runoff = FALSE, ggtheme = "light")
```

![rain amount][rain]

## Wind

There are three types of plots available for wind data in `clifro`. The default
is to plot a windrose displaying wind speed and directions of the full time 
series at each station. The `windrose` function in `clifro` is also available 
for the user to plot their own directional data. The other two optional plots 
for wind data in `clifro` are the wind speed and wind direction plots. These 
plots display wind speed and direction patterns through time, adding valuable 
temporal information that is not portrayed in the windroses.

### Windrose

```{r,eval=FALSE}
# Defaults to windrose
plot(cf.datalist, which = 4)
```

![windrose][windrose]

### Wind Speeds

```{r,eval=FALSE}
# Plot the wind speeds through time
plot(cf.datalist, which = 4, wind_plot = "speed", ggtheme = "classic")
```

![wind speed][windSpeed]

### Wind Directions

```{r,eval=FALSE}
# Plot wind direction contours
plot(cf.datalist, which = 4, wind_plot = "direction", contours = 10, ggtheme = "light")
```

![wind direction][windDir]

# Export

```{r,eval=FALSE}
# Export the data as separate CSV files to the current working directory
getwd()
for (i in seq_along(cf.datalist))
  write.csv(cf.datalist[i], 
            file = tempfile(paste0(cf.datalist[i]@dt_name, "_"), 
                            tmpdir = normalizePath("."), 
                            fileext = ".csv"))
```


# Summary

The primary aim of this package is to make the substantial amount of climate 
data residing within the National Climate Database more accessible and easier
to work with. The `clifro` package has many advantages over using the CliFlo 
web portal including conducting searches much more efficiently, examining the
spatial extent of the stations and enabling high quality plots to aid the data
exploration and analysis stage.

[chooseStation]: choose-station.html
[chooseDatatype]: choose-datatype.html
[clifrostation]: cfStation.html

[mslAtmosPress]: figures/atmospress.png "MSL Atmospheric Pressure"
[mslAtmosPress2]: figures/atmospress2.png "Improved MSL Atmospheric Pressure"
[temperature]: figures/temperature.png "Temperature Extremes"
[rainRunoff]: figures/rainRunoff.png "Rain with Soil Deficit and Runoff"
[rain]: figures/rain.png "Rain Amount"
[windrose]: figures/windrose.png "Windrose"
[windSpeed]: figures/windspeed.png "Wind Speed"
[windDir]: figures/winddir.png "Wind Direction"